How is experience transformed into memory?

### 2018-08-28 paper/main.tex

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Supplementary Figure},fileext=lof]{suppfigure}

\title{How is experience transformed into memory?}
% \title{Observing the transformation of dynamic experiences into memory}
\author{Andrew C. Heusser, Paxton C. Fitzpatrick \& Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that, despite the complexities of real-world experiences, they unfold in a structured and often predictable manner. This structure gives each of our experiences (and thus our memories for those experiences) a unique `shape' with respect to how it unfolds in the relevant representational space.  However, temporal structure is typically absent (or severely impoverished) in traditional trial-based  experiments. As a consequence, our understanding of how the temporal structure (i.e., the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the video to participants' later recountings of the narrative. Using a novel approach that leverages topic modeling and Hidden Markov Models, we found that, despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by nearly every participant. Our findings suggest that assessing the degree of match between the shape of an experience and the subsequent retelling of that experience provides a much more nuanced view (compared with traditional methods such as comparing the proportion of remembered details) into what it means to remember. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a binary operation: either a studied item is recalled or it isn't. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' an experience or a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) concerns a much different set of criteria.  First, real-world recall is continuous, rather than binary.  Second, the specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Third, one might remember the gist or essence of an experience but forget (or neglect to recount) particular details.  Or different people who share an experience might recount the experience with a similar level of detail, but the specific details that were remembered might vary across people.  Which aspects of those recollections should be considered fundamental to determining to whast extent or how reliably an experience has been remembered?

%JRM STOPPED INPUTTING EDITS HERE

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e., how the contents relate/change over time).  For instance, naturalistic experiences are autocorrelated in space and time on short timescales. In other words, the contents of an experience often change gradually and are therefore similar from moment to moment. As an example, many of us spend our mornings/evenings at home and much of our days at work (e.g.\ in an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) within an experience. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall the isolated elements of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. These recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical to understanding how our memory systems encode, store and retrieve our life experiences.

While our experiences often change gradually, they can also sometimes change suddenly (e.g.\ walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\subsection{Results}

@@

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}A).

@@

\subsubsection{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}C,D).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most thematically similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}C,D).

@@

\subsubsection{The trajectory of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space using the UMAP dimensionality reduction algorithm ~\citep{McInHeal18}, where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}A).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space.'' For each point on the grid, we drew a circle (radius=.5) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test ($p<.001$, corrected at $p<.05$ using the permutation procedure described in Methods). Notably, the video events fell within the path formed by the recall events.  Thus, participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}B), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

%NOTE: it feels like we need some more supporting evidence that the shape of the video and the recall trajectories are related.  one possibility would be to add the analysis where i compute the distance between each recall event and the line formed by each video transition per subject. the result is that hand annotated memory correlates to this distance, suggesting that subjects who better capture the shape also have better memory on average.

\subsubsection{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} \textbf{A.} The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  \textbf{B.} The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

@@

While each participant remembered roughly the same sequence of events, there were subtle idiosyncracies in the way each individual described the events. For example, the event-by-event similarity structure (i.e. the off-diagonal values of the recall correlation matrices) differed across individuals (Supp. Fig.~\ref{fig:corrmats}). We hypothesized that idiosyncracies in the recall dynamics across subjects might be related to the temporal structure of brain networks representing the events during video viewing. In this analysis, we identified brain areas whose responses at the time of viewing exhibited the same (segmented) idiosyncratic dynamics that later appeared in that participants’ recalls. Using a procedure similar to the approach above (see Methods for details), we searched for regions whose correlational structure during viewing was related to the idiosyncratic structure of the participant's recall model. The resulting map suggests that the correlational structure of a network of brain regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe mirror the idiosyncratic structure of participants' recalls (Fig.~\ref{fig:brainz}).

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} \textbf{A.} Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants \textbf{B.} Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. \textbf{C.} Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are individual recall events across all participants, where the colors correspond to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

\subsubsection{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli and memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis (and visualization) of the language used when participants recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the ``top'' words for the most and least memorable scenes.

@@

\subsubsection{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than was previously possible. We quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants ($p<.05$, see Supplemental Materials Section 1 for details).

@@

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes recollection of an experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the video participants viewed by measuring the match between the shapes of the video's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory -- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

@@

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g.\ one that describes the video in the same way as participants recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g.\ participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to such details while maintaining flexibility.

@@

\subsection{Participants and Experimental Design}
Participants ($n=17$) viewed the first 50 minutes of ``A Study in Pink'', an episode of the BBC series, \textit{Sherlock}. Immediately upon completion of the video, participants were instructed to (verbally) recount the events in the \textit{Sherlock} episode in their original order and in as much detail as possible. During the entire experiment, participants were in an fMRI scanner. For comprehensive details of the experimental procedures, please refer to \cite{ChenEtal17}.
%NOTE: is this sufficient or should i provide more detail?

\subsubsection{Fitting the topic model to the video text and recall transcripts}
A topic model was used to estimate the most likely mixture of topics for a given sample of text. First, the video was manually segmented into 1000 scenes, and a collection of descriptive features was manually transcribed. For each scene, we considered the following features: narrative details (a sentence or two describing what happened in that scene), whether the scene was indoor or outdoor, name of all the characters in the scene, name of the character in focus, name of the character speaking, location, camera angle, music presence, and text on the screen. We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene. We then transformed the text descriptions into overlapping windows of 50 scene segments. For example, the first text sample comprised the text from the first 50 segments, the 2nd comprised the text from n+1:n+51, and so on. We trained our model using these overlapping text samples with scikit-learn's (version 0.19.1) `CountVectorizer' and `LatentDirichletAllocation' classes \citep{PedrEtal11} implemented using our high-dimensional visualization/analysis software, Hypertools \citep{HeusEtal17a}.  First, the text was transformed into a vector of word counts (after removing English stopwords). This gave a word count vector for each scene in the video.  Then, the word count vectors were used to fit a topic model (topics=100, method=`batch'). We transformed the text descriptions using the model resulting in a scenes (1000) by topics (100) matrix. The scene descriptions often spanned multiple timepoints (i.e., TRs). To account for this, we expanded the video model by copying the rows of the model for as many timepoints that the scene description spanned. After this expansion, the shape of the model was the length of the duration of the video (1976 TRs).

@@

\subsubsection{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g.\ hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

\subsubsection{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events, see Supp. Fig.~\ref{fig:corrmats}). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events ~\citep{Rabi89, BaldEtal17}. To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17, Brainiak}.

>>>>

% !TEX TS-program = pdflatex
\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Supplementary Figure},fileext=lof]{suppfigure}

\title{How is experience transformed into memory?}
\author{Andrew C. Heusser, Paxton C. Fitzpatrick, and Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
  How our experiences unfold over time define unique \textit{thought
    trajectories} through the relevant representational spaces.
  Within this geometric framework, one can compare the shape of the
  trajectory formed by an experience to that defined by our later
  remembering of that experience.  We propose a framework for mapping
  naturalistic experiences onto geometric spaces that characterize how
  they unfold over time.  New insights emerge when we apply this
  approach to a naturalistic memory experiment which had participants
  view and recount a video.  We found that the shapes of the
  trajectories formed by participants' recountings were all highly
  similar to that of the original video, but participants differed in
  the level of detail they remembered.  We also identified a network
  of brain structures that are sensitive to the ``shapes'' of our
  ongoing experiences, and an overlapping network that is sensitive to
  how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional
episodic memory experiments \citep[e.g., list-learning or trial-based
experiments;][]{Murd62a, Kaha96}, remembering is often cast as a
discrete and binary operation: each studied item may be separated from
the rest of one's experiences, and that item may be labeled as having
been recalled versus forgotten. More nuanced studies might incorporate
self-reported confidence measures as a proxy for memory strength, or
ask participants to discriminate between ``recollecting'' the
(contextual) details of an experience or having a general feeling of
``familiarity'' \citep{Yone02}. However, characterizing and evaluating
memory in more realistic contexts (e.g., recounting a recent
experience to a friend) is fundamentally different.  First, real-world
recall is continuous, rather than discrete.  Removing an event from
the context in which it occurrs often subtantially changes its
meaning.  Second, the specific words used to describe an experience
have very little bearing on whether the experience is considered to
have been ``remembered.''  Asking whether the rememberer has precisely
reproduced a specific set of words to describe a given experience is
nearly orthogonal to whether they were actually able to remember it.
Third, one might remember the gist or essence of an experience but
forget (or neglect to recount) particular details.  Capturing the gist
of what happened is typically the ``point'' of recounting a memory to
a listener but, depending on the circumstances, accurate recall of any
specific detail may be irrelevant.

% JRM REWRITES STOPPED HERE

%(JRM: started on the next paragraph...)
Human memory studies typically assess memory for the elements within
an experience, such as the people, places and things
encountered. However, real life experiences can also be described in
terms of their temporal structure (i.e., how the contents
relate/change over time).  For instance, naturalistic experiences are
autocorrelated in space and time on short timescales. In other words,
the contents of an experience often change gradually and are therefore
similar from moment to moment. As an example, many of us spend our
mornings/evenings at home and much of our days at work (e.g., in an
office). While the specific happenings within these contexts differ,
they are often highly correlated (and thus predictable) within an
experience. Furthermore, our experiences can also be characterized by
longer timescale correlations. For example, consider returning to your
office after taking a lunch break. The spatial contexts before and
after lunch are highly similar, but separated in time. Thus, while
evaluating the ability to recall the isolated elements of an
experience has been a fruitful way to learn about human memory,
another important dimension to consider is the temporal structure of
the remembered experience, and how the temporal structure of the
memory relates to the original experience. These recurrent and
gradually changing temporal dynamics are not typically present in
traditional memory studies, but are likely critical to understanding
how our memory systems encode, store and retrieve our life
experiences.

While our experiences often change gradually, they can also sometimes change suddenly (e.g., walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

\subsection*{Results}

@@

\subsubsection*{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters present, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}A, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}C).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (a few sentences summarizing the dialogue and specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}A).

@@

\subsubsection*{Segmenting the video and recall models into ``events''}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}C,D).  We hypothesized that this structure might arise from transient stability in the "theme" of both the video and of participants' memory for the video. We hypothesized that specific events described by a participant could be ``matched'' to specific video events by computing the similarity (i.e., correlation) between their topic vectors. In other words, the video event that is most thematically similar to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e., states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}C,D).

@@

\subsubsection*{The trajectory of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches cannot (and were not designed to) capture the rich temporal dynamics inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space using the UMAP dimensionality reduction algorithm ~\citep{McInHeal18}, where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}A).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space.'' For each point on the grid, we drew a circle (radius=.5) around the point and grouped together all recall transitions that intersected the circle (across participants). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test ($p<.001$, corrected at $p<.05$ using the permutation procedure described in Methods). Notably, the video events fell within the path formed by the recall events.  Thus, participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}B), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

%NOTE: it feels like we need some more supporting evidence that the shape of the video and the recall trajectories are related.  one possibility would be to add the analysis where i compute the distance between each recall event and the line formed by each video transition per participant. the result is that hand annotated memory correlates to this distance, suggesting that participants who better capture the shape also have better memory on average.

\subsubsection*{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} \textbf{A.} The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  \textbf{B.} The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

@@

While each participant remembered roughly the same sequence of events, there were subtle idiosyncracies in the way each individual described the events. For example, the event-by-event similarity structure (i.e., the off-diagonal values of the recall correlation matrices) differed across individuals (Supp. Fig.~\ref{fig:corrmats}). We hypothesized that idiosyncracies in the recall dynamics across participants might be related to the temporal structure of brain networks representing the events during video viewing. In this analysis, we identified brain areas whose responses at the time of viewing exhibited the same (segmented) idiosyncratic dynamics that later appeared in that participants’ recalls. Using a procedure similar to the approach above (see Methods for details), we searched for regions whose correlational structure during viewing was related to the idiosyncratic structure of the participant's recall model. The resulting map suggests that the correlational structure of a network of brain regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe mirror the idiosyncratic structure of participants' recalls (Fig.~\ref{fig:brainz}).

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} \textbf{A.} Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants \textbf{B.} Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g., the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. \textbf{C.} Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are individual recall events across all participants, where the colors correspond to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

\subsubsection*{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli and memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis (and visualization) of the language used when participants recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the ``top'' words for the most and least memorable scenes.

@@

\subsubsection*{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than was previously possible. We quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants ($p<.05$, see Supplemental Materials Section 1 for details).

@@

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes recollection of an experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the video participants viewed by measuring the match between the shapes of the video's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory -- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g., percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

@@

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g., one that describes the video in the same way as participants recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g., participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to such details while maintaining flexibility.

@@

\subsection*{Participants and Experimental Design}
Participants ($n=17$) viewed the first 50 minutes of ``A Study in Pink'', an episode of the BBC series, \textit{Sherlock}. Immediately upon completion of the video, participants were instructed to (verbally) recount the events in the \textit{Sherlock} episode in their original order and in as much detail as possible. During the entire experiment, participants were in an fMRI scanner. For comprehensive details of the experimental procedures, please refer to \cite{ChenEtal17}.
%NOTE: is this sufficient or should i provide more detail?

\subsubsection*{Fitting the topic model to the video text and recall transcripts}
A topic model was used to estimate the most likely mixture of topics for a given sample of text. First, the video was manually segmented into 1000 scenes, and a collection of descriptive features was manually transcribed. For each scene, we considered the following features: narrative details (a sentence or two describing what happened in that scene), whether the scene was indoor or outdoor, name of all the characters in the scene, name of the character in focus, name of the character speaking, location, camera angle, music presence, and text on the screen. We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene. We then transformed the text descriptions into overlapping windows of 50 scene segments. For example, the first text sample comprised the text from the first 50 segments, the 2nd comprised the text from n+1:n+51, and so on. We trained our model using these overlapping text samples with scikit-learn's (version 0.19.1) `CountVectorizer' and `LatentDirichletAllocation' classes \citep{PedrEtal11} implemented using our high-dimensional visualization/analysis software, Hypertools \citep{HeusEtal18}.  First, the text was transformed into a vector of word counts (after removing English stopwords). This gave a word count vector for each scene in the video.  Then, the word count vectors were used to fit a topic model (topics=100, method=`batch'). We transformed the text descriptions using the model resulting in a scenes (1000) by topics (100) matrix. The scene descriptions often spanned multiple timepoints (i.e., TRs). To account for this, we expanded the video model by copying the rows of the model for as many timepoints that the scene description spanned. After this expansion, the shape of the model was the length of the duration of the video (1976 TRs).

@@

\subsubsection*{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g., hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.
%NOTE: I did run this analysis, but before that i arbitrarily selected some parameters that ended up being very close to the best.  The reason I bring this up is that the correlations between some of the memory metrics (k, precision, distinctiveness) and hand annotated accuracy could be biased because we searched for the optimal parameters. Its not `the same' measure, but they are related to each other, so we'll have to be extra careful not to over interpret the correlations in this study.

\subsubsection*{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events, see Supp. Fig.~\ref{fig:corrmats}). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events ~\citep{Rabi89, BaldEtal17}. To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17, Brainiak}.
